---
title: "envfit_analysis"
output: pdf_document
author: "Jacob Westaway"
date: "Last updated on `r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/Jacob/Desktop/Jacob_Uni/Data/SCN_vs_NICU_probiotic_study')
load(file = "Rmd.RData")
```

# About

EnvFit analysis (Vegan package) was used to explore the effect of covariates on microbial community composition between groups (beta diversity), to determine if probiotic treatment has an effect. The EnvFit function was applied to an ordination using NMDS that was based on a Bray-Curtis dissimilarity matrix calculated from data normalised with Total Sum Scaling. The significance was based on 10,000 permutations and was transformed based on the Benjamini-Hochberg (BH) procedure.

# Load packages

```{r, warning=F, message=F, results='hide'}
sapply(c("DESeq2", "phyloseq", "dplyr", "ggplot2", "grid", 
         "gridExtra", "ggpubr", "sjPlot", "pheatmap", "tidyverse", "vegan"), 
          require, character.only = TRUE)
```

# Explore clustering of variables with PCoA and Bray-Curtis

```{r, fig.cap="PCoA plot of Bray-Curtis distances coloured by probiotic group.", warning=F, message=F}
# define function
ordination_plots <- function(filtered_ps, variable, vis_method, dist_method){
# ordinate
ps_ordination <- ordinate(filtered_ps, method = vis_method, distance = dist_method)
# get eignenvalues
evals <- ps_ordination$values$Eigenvalues
# generate plot
plot_ordination(filtered_ps, ps_ordination, color = variable, 
  title = "PCoA (Bray-Curtis)") +
  labs(col = variable) +
  coord_fixed(sqrt(evals[2] / evals[1])) +
  geom_point(size = 3) +
  stat_ellipse(typre = "norm", linetype = 2) +
  scale_color_hue(labels = c("Probiotic-treated", "Non-treated"))
 }

ordination_plots(ps2.TSS, "Primary_Group", "PCoA", "bray")
```

# envfit analyis 

```{r,warning=F,message=F,eval=F}
ordination_distance <- ordinate(
    physeq = ps2.TSS, 
    method = "NMDS",
    distance = "bray",
    formula = ~ Primary_Group + Feeding_Type + NEC + Sepsis + Mode_of_Delivery + 
      Neonatal_Antibiotics + Chorioamnionitis + Preeclampsia + ROP + Batch + 
      Gestational_Age_at_Birth + Diabetes + Antenatal_Antibiotics)

envdat.phy <- ps2.TSS %>% 
  sample_data() %>%
  unclass() %>% 
  as.data.frame() %>% 
  select(Primary_Group, Feeding_Type, NEC, Sepsis, Mode_of_Delivery, 
         Neonatal_Antibiotics, Chorioamnionitis, Preeclampsia, ROP, Batch, 
         Gestational_Age_at_Birth, Diabetes, Antenatal_Antibiotics) %>% 
  centre_and_scale()

set.seed(1)

envfit_object <- envfit(ordination_distance, envdat.phy, permutations = 10000) 
``` 


# Adjust p values

```{r,warning=F,message=F,eval=F}
p.adjust.envfit <- function (x, method = 'bonferroni', n){
  x.new <- x
  if (!is.null (x$vectors)) pval.vectors <- x$vectors$pvals else pval.vectors <- NULL
  if (!is.null (x$factors)) pval.factors <- x$factors$pvals else pval.factors <- NULL
  if (missing (n)) n <- length (pval.vectors) + length (pval.factors)
  if (!is.null (x$vectors)) x.new$vectors$pvals <- p.adjust (x$vectors$pvals, 
                                                             method = method, n = n)
  if (!is.null (x$factors)) x.new$factors$pvals <- p.adjust (x$factors$pvals, 
                                                             method = method, n = n)
  cat ('Adjustment of significance by', method, 'method')
  return (x.new)
}

set.seed(1)

envfit(ordination_distance, envdat.phy, permutations = 10000) %>% 
  p.adjust.envfit(., method = "BH") 
```

# Plot model

```{r,warning=F,message=F}
envfit_object$vectors$r %>% 
  as.data.frame() %>% 
  rownames_to_column("Variable") %>%  
  rename("r2" = ".") %>%  
  left_join(
    (envfit_object$vectors$pvals %>% 
  as.data.frame() %>% 
  add_column(Variable = "Gestational_Age_at_Birth") %>% 
  rename("p" = "."))) %>% 
  rbind(
envfit_object$factors$r %>% 
  as.data.frame() %>% 
  rownames_to_column("Variable") %>%  
  rename("r2" = ".") %>% 
  rows_update(., tibble(Variable = "Sepsis", r2 = 0.1)) %>% # changed to 0.1 for plot 
  left_join(
    envfit_object$factors$pvals %>% 
  as.data.frame() %>% 
  rownames_to_column("Variable") %>%  
  rename("p" = "."))
  ) %>% 
  ggplot(aes(x = r2, y = Variable, fill = p)) +
  geom_col() +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10), limits = c(0, 0.1)) +
  theme(axis.line =  element_line(size = .5), panel.background = element_blank()) + 
  scale_fill_continuous(type = "viridis") +
  scale_y_discrete(labels = c("Primary_Group" = "Probiotic group",
                              "Neonatal_Antibiotics" = "Neonatal antibiotics",
                              "Mode_of_Delivery" = "Mode of delivery",
                              "Feeding_Type" = "Diet",
                              "Birth_Weight" = "Birthweight",
                              "Gestational_Age_at_Birth" = "Gestation at birth",
                              "Antenatal_Antibiotics" = "Antenatal antibiotics"))
```

# Calculate major contributors (based on adonis)

```{r, warning=F, message=F}
major_contributors <- function(ps2.TSS, variable){
# perform permanova  
ps_otu <- data.frame(otu_table(ps2.TSS))
ps_metadata <- data.frame(sample_data(ps2.TSS))
permanova <- adonis(ps_otu ~Primary_Group + Feeding_Type + NEC + Sepsis + 
                    Mode_of_Delivery + Neonatal_Antibiotics + Chorioamnionitis +
                    Preeclampsia + ROP + Batch + Gestational_Age_at_Birth + Diabetes + 
                    Antenatal_Antibiotics, data = ps_metadata, method = "bray")  

# coefficients
coef <- coefficients(permanova)[paste0(variable, "1"),]
top.coef <- coef[rev(order(abs(coef)))[1:20]]

genus_contributors <- tax_table(ps2.TSS) %>% 
    unclass() %>% 
    as.data.frame() %>% 
    select("Genus") %>% 
    rownames_to_column(var = "ASV") %>% 
    right_join((as.data.frame(top.coef) %>%
    rownames_to_column(var = "ASV"))) %>% 
    select(!"ASV") %>% 
    arrange(top.coef)

return(genus_contributors)
}

major_contributors(ps2.TSS, "Primary_Group") %>% 
  filter(top.coef > 0) %>% 
  group_by(Genus) %>% 
  summarise("top.coef" = mean(top.coef)) %>% # ASVs from same genus 
  rbind(
    major_contributors(ps2.TSS, "Primary_Group") %>% 
  filter(top.coef < 0) %>% 
  group_by(Genus) %>% 
  summarise("top.coef" = mean(top.coef))
  ) %>%  
  ggplot(aes(x = top.coef, y = Genus, fill = top.coef)) +
  geom_col() +
  theme(axis.line =  element_line(size = .5), panel.background = element_blank()) +
  geom_vline(xintercept = 0.0, linetype = 1, size = .5) + 
  scale_fill_continuous(type = "viridis", name = "Coefficient") +
  theme(axis.title.x = element_blank())
```