---
title: "Deleted"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 - Define a function to extract the results.
 - Extract the results, order by p value, selects significant (<0.05) results, binds this data to the *tax_table* from the *phyloseq* object to get the taxonomic information, and then select and order the desired columns.

```{r, warning=F, message=F}
# function for liklihood ratio test
get_deseq_res_lrt <- function(deseq_object){
  res = results(deseq_object)
  res = res[order(res$padj, na.last = NA), ]
  sigtab = res[(res$padj < 0.05), ] 
  sigtab = cbind(as(sigtab, "data.frame"), 
          as(tax_table(ps3)[rownames(sigtab), ], "matrix"))
  sigtab %>%
  arrange(padj) %>%
  select("log2FoldChange", "lfcSE", "padj", "Genus") 
}
# function for Walds test and continuous variables
get_deseq_res_cont <- function(deseq_object, contrast_variable){
  res = results(deseq_object, name = contrast_variable)
  res = res[order(res$padj, na.last = NA), ]
  sigtab = res[(res$padj < 0.05), ] 
  sigtab = cbind(as(sigtab, "data.frame"), 
          as(tax_table(ps3)[rownames(sigtab), ], "matrix"))
  sigtab %>%
  arrange(padj) %>%
  select("log2FoldChange", "lfcSE", "padj", "Genus") 
}
# function for Walds test and categorical variables
get_deseq_res_cat <- function(desq_object, contrast_variable, level1, level2){
  res = results(desq_object, contrast = c(contrast_variable, level1, level2))
  res = res[order(res$padj, na.last = NA), ]
  alpha = 0.05
  sigtab = res[(res$padj < alpha), ] 
  sigtab = cbind(as(sigtab, "data.frame"), 
    as(tax_table(ps3)[rownames(sigtab), ], "matrix"))
  sigtab %>%
  arrange(padj) %>%
  select("log2FoldChange", "lfcSE", "padj", "Genus") %>%
  add_column(Variable = paste0(contrast_variable, level1)) # label the base level
}
```